---
title: "Forecasting Seoul's hourly bike demand"
subtitle: "A leakage-safe operational case study with interpretable benchmarks"
author: "Miguelangel Palma Rojas"
date: last-modified
lang: en
description: "An end-to-end forecasting workflow for bicycle rebalancing, evaluated on a strictly future holdout."
---

```{r}
#| label: setup
#| include: false
source(here::here("R", "analysis.R"))
library(knitr)
bikes <- read_bike_snapshot()
results <- fit_bike_models(bikes, seed = 2026)
profile <- bike_data_profile(bikes)
leader <- results$metrics |> slice_min(rmse, n = 1)
baseline_rmse <- results$metrics |> filter(model == "Calendar baseline") |> pull(rmse)
improvement <- 100 * (baseline_rmse - leader$rmse) / baseline_rmse
```

Hourly demand is not just a prediction problem. It is an operating decision: how many bicycles should be available, where should the team expect pressure, and how much uncertainty remains at commuter peaks?

::: {.metric-grid}
::: {.metric-card}
<strong>`r leader$model`</strong>
<span>Best final model</span>
:::
::: {.metric-card}
<strong>`r sprintf("%.1f", leader$rmse)` bikes</strong>
<span>Holdout RMSE</span>
:::
::: {.metric-card}
<strong>`r sprintf("%.1f%%", improvement)`</strong>
<span>RMSE gain vs calendar baseline</span>
:::
::: {.metric-card}
<strong>`r sprintf("%.0f", leader$peak_mae)` bikes</strong>
<span>Peak-demand MAE</span>
:::
:::

::: {.callout-important}
## Decision takeaway
Use the winning model as a planning signal, not an autonomous dispatch rule. It materially improves the calendar benchmark, but its negative bias and larger peak error call for an operational safety buffer.
:::

## Data contract and quality

The committed snapshot contains `r format(profile$observations, big.mark = ",")` hourly observations from `r profile$first_date` through `r profile$last_date`. The pipeline validates schema, completeness, physical ranges, nonnegative rentals, and unique date-hour keys before modeling.

```{r}
#| label: inspect-data
bikes <- read_bike_snapshot()
bike_data_profile(bikes) |>
  kable(digits = 1, caption = "Validated snapshot profile")
```

```{r}
#| label: demand-rhythm
#| fig-cap: "Average operating-day rentals by weekday and hour."
plot_demand_heatmap(bikes, "en")
```

## Forecast design

The last 20% of complete dates is untouched until the final comparison. Hyperparameters are selected with expanding temporal windows; every validation window follows its training window and a one-day gap reduces boundary leakage.

```{r}
#| label: temporal-split
model_data <- prepare_modeling_data(bikes)
partition <- temporal_partition(model_data, test_fraction = 0.20)
resamples <- temporal_resamples(
  partition$training,
  folds = 4,
  assessment_days = 28,
  gap_days = 1
)
temporal_window_summary(list(resamples = resamples)) |>
  kable(caption = "Expanding validation windows")
```

The recipe treats date as an identifier, encodes categorical calendar signals, removes zero-variance and linearly dependent columns, and normalizes numeric predictors. Four candidates make the result auditable: a weekday-hour benchmark, ordinary linear regression, Elastic Net, and random forest.

Independent fixed seeds govern each search grid and the random-forest engine, so both language editions reproduce the same result.

```{r}
#| label: modeling-recipe
recipe <- bike_recipe(partition$training)

forest_spec <- rand_forest(
  mtry = tune(), min_n = tune(), trees = 300
) |>
  set_engine("ranger", importance = "permutation") |>
  set_mode("regression")

workflow() |>
  add_recipe(recipe) |>
  add_model(forest_spec)
```

## Final holdout results

```{r}
#| label: metrics-table
results$metrics |>
  mutate(across(where(is.numeric), ~ round(.x, 1))) |>
  kable(caption = "Operational metrics on unseen future dates")
```

```{r}
#| label: model-comparison
#| fig-cap: "Lower RMSE is better; evaluation uses only the final future holdout."
plot_model_metrics(results$metrics, "en")
```

RMSE penalizes large misses, MAE describes the typical absolute miss, WAPE expresses total error relative to observed demand, bias shows systematic over- or under-forecasting, and peak MAE isolates the busiest 10% of training demand.

```{r}
#| label: holdout-trace
#| fig-cap: "Observed demand and the winning model over the final seven days."
plot_holdout_series(results, language = "en")
```

## Operational diagnostics

```{r}
#| label: peak-errors
#| fig-cap: "Positive residuals indicate under-forecasting."
plot_peak_residuals(results, "en")
```

```{r}
#| label: feature-importance
#| fig-cap: "Permutation importance describes predictive contribution, not causality."
plot_variable_importance(results$importance, "en")
```

Temperature, trend, moisture conditions, solar radiation, and commute-hour indicators carry much of the signal. This does not make them causal levers; it makes them useful forecasting inputs.

## From forecast to action

- Refresh the forecast before each operating shift with the latest weather outlook.
- Convert expected rentals into station-level targets only after adding station inventory and trip-flow data.
- Add a peak buffer when the cost of stockouts exceeds the cost of carrying idle bicycles.
- Monitor WAPE, bias, and peak MAE by season; retrain when performance drifts beyond agreed thresholds.
- Keep closures outside the demand model and handle them with an explicit service-status rule.

The dataset covers one year and has no station, capacity, pricing, event, or forecast-weather fields. The result demonstrates sound model evaluation, but it is not yet a production rebalancing system. See the [methodology](docs/methodology.md), [model card](docs/model-card.md), and [data dictionary](docs/data-dictionary.md) for the complete audit trail.
