---
title: "Forecasting Seoul's hourly bike demand"
subtitle: "A self-contained tidyverse and tidymodels case study"
author: "Miguelangel Palma Rojas"
date: last-modified
lang: en
description: "Visible, end-to-end forecasting code for bicycle operations, evaluated on a strictly future holdout."
---

Hourly demand is not just a prediction problem. It is an operating decision: how many bicycles should be available, where should the team expect pressure, and how much uncertainty remains at commuter peaks?

This notebook is intentionally self-contained. Every transformation, split, model, metric, and chart used below is written here and executed from top to bottom.

## Packages and reproducibility

```{r}
#| label: packages
library(tidyverse)
library(tidymodels)
library(ranger)
library(knitr)

tidymodels_prefer()
set.seed(2026)
```

## Data contract and quality

The governed snapshot contains one row per date and hour. Before modeling, the notebook verifies the complete schema, missingness, physical ranges, nonnegative rentals, and unique date-hour keys.

```{r}
#| label: read-and-validate
required_columns <- c(
  "date", "rented_bike_count", "hour", "temperature_c",
  "humidity_pct", "wind_speed_ms", "visibility_10m", "dew_point_c",
  "solar_radiation_mj_m2", "rainfall_mm", "snowfall_cm", "season",
  "holiday", "functioning_day", "weekday", "is_weekend"
)

bikes <- read_csv(
  "data/seoul_bike_snapshot.csv",
  show_col_types = FALSE,
  col_types = cols(date = col_date())
) |>
  mutate(
    weekday = factor(
      weekday,
      levels = c(
        "Monday", "Tuesday", "Wednesday", "Thursday",
        "Friday", "Saturday", "Sunday"
      )
    ),
    across(c(season, holiday, functioning_day, is_weekend), factor)
  ) |>
  arrange(date, hour)

stopifnot(
  setequal(names(bikes), required_columns),
  !anyNA(bikes),
  all(bikes$rented_bike_count >= 0),
  all(bikes$hour %in% 0:23),
  between(min(bikes$humidity_pct), 0, 100),
  between(max(bikes$humidity_pct), 0, 100),
  nrow(distinct(bikes, date, hour)) == nrow(bikes)
)

data_profile <- bikes |>
  summarise(
    observations = n(),
    first_date = min(date),
    last_date = max(date),
    operating_days = n_distinct(date[functioning_day == "Yes"]),
    closure_hours = sum(functioning_day == "No"),
    average_hourly_demand = mean(rented_bike_count),
    peak_hourly_demand = max(rented_bike_count),
    missing_cells = sum(is.na(pick(everything())))
  )

data_profile |>
  kable(digits = 1, caption = "Validated snapshot profile")
```

```{r}
#| label: demand-rhythm
#| fig-cap: "Average operating-day rentals by weekday and hour."
bikes |>
  filter(functioning_day == "Yes") |>
  summarise(
    average_demand = mean(rented_bike_count),
    .by = c(weekday, hour)
  ) |>
  ggplot(aes(hour, weekday, fill = average_demand)) +
  geom_tile() +
  scale_fill_viridis_c(option = "C") +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(
    title = "Commute peaks define the operating rhythm",
    x = "Hour", y = "Weekday", fill = "Average rentals"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), panel.grid = element_blank())
```

## Feature engineering and future holdout

Closures are handled as a service-status rule, not learned as zero demand. The latest 20% of complete dates is reserved untouched for final evaluation.

```{r}
#| label: feature-engineering
origin_date <- min(bikes$date)

model_data <- bikes |>
  filter(functioning_day == "Yes") |>
  transmute(
    date,
    rented_bike_count,
    hour = factor(hour, levels = 0:23),
    weekday,
    month = factor(format(date, "%m"), levels = sprintf("%02d", 1:12)),
    season,
    holiday,
    is_weekend,
    trend_day = as.numeric(date - origin_date),
    temperature_c,
    humidity_pct,
    wind_speed_ms,
    visibility_10m,
    dew_point_c,
    solar_radiation_mj_m2,
    rainfall_mm,
    snowfall_cm
  ) |>
  arrange(date, hour)

test_days <- ceiling(n_distinct(model_data$date) * 0.20)

first_test_date <- model_data |>
  distinct(date) |>
  arrange(desc(date)) |>
  slice_head(n = test_days) |>
  summarise(first_test_date = min(date)) |>
  pull(first_test_date)

training_data <- model_data |>
  filter(date < first_test_date)

testing_data <- model_data |>
  filter(date >= first_test_date)

tibble(
  sample = c("Training", "Future holdout"),
  first_date = c(min(training_data$date), min(testing_data$date)),
  last_date = c(max(training_data$date), max(testing_data$date)),
  observations = c(nrow(training_data), nrow(testing_data))
) |>
  kable(caption = "Chronological final split")
```

## Expanding temporal validation

Hyperparameters are selected only inside the training period. Each 28-day validation window follows an expanding training window and a one-day gap.

```{r}
#| label: temporal-resampling
training_dates <- training_data |>
  distinct(date) |>
  arrange(date) |>
  pull(date)

assessment_days <- 28L
gap_days <- 1L
analysis_ends <- round(seq(
  floor(length(training_dates) * 0.45),
  length(training_dates) - assessment_days - gap_days,
  length.out = 4
)) |>
  as.integer()

assessment_starts <- analysis_ends + gap_days + 1L
assessment_ends <- assessment_starts + assessment_days - 1L

split_01 <- make_splits(
  list(
    analysis = which(training_data$date %in% training_dates[seq_len(analysis_ends[[1]])]),
    assessment = which(training_data$date %in% training_dates[assessment_starts[[1]]:assessment_ends[[1]]])
  ),
  training_data
)

split_02 <- make_splits(
  list(
    analysis = which(training_data$date %in% training_dates[seq_len(analysis_ends[[2]])]),
    assessment = which(training_data$date %in% training_dates[assessment_starts[[2]]:assessment_ends[[2]]])
  ),
  training_data
)

split_03 <- make_splits(
  list(
    analysis = which(training_data$date %in% training_dates[seq_len(analysis_ends[[3]])]),
    assessment = which(training_data$date %in% training_dates[assessment_starts[[3]]:assessment_ends[[3]]])
  ),
  training_data
)

split_04 <- make_splits(
  list(
    analysis = which(training_data$date %in% training_dates[seq_len(analysis_ends[[4]])]),
    assessment = which(training_data$date %in% training_dates[assessment_starts[[4]]:assessment_ends[[4]]])
  ),
  training_data
)

temporal_splits <- list(split_01, split_02, split_03, split_04)

temporal_resamples <- manual_rset(
  temporal_splits,
  paste0("Window", str_pad(seq_along(temporal_splits), 2, pad = "0"))
)

window_summary <- tibble(
  window = temporal_resamples$id,
  training_start = c(
    min(analysis(split_01)$date), min(analysis(split_02)$date),
    min(analysis(split_03)$date), min(analysis(split_04)$date)
  ),
  training_end = c(
    max(analysis(split_01)$date), max(analysis(split_02)$date),
    max(analysis(split_03)$date), max(analysis(split_04)$date)
  ),
  validation_start = c(
    min(assessment(split_01)$date), min(assessment(split_02)$date),
    min(assessment(split_03)$date), min(assessment(split_04)$date)
  ),
  validation_end = c(
    max(assessment(split_01)$date), max(assessment(split_02)$date),
    max(assessment(split_03)$date), max(assessment(split_04)$date)
  ),
  training_rows = c(
    nrow(analysis(split_01)), nrow(analysis(split_02)),
    nrow(analysis(split_03)), nrow(analysis(split_04))
  ),
  validation_rows = c(
    nrow(assessment(split_01)), nrow(assessment(split_02)),
    nrow(assessment(split_03)), nrow(assessment(split_04))
  )
)

window_summary |>
  kable(caption = "Expanding validation windows")
```

## Preprocessing and model specifications

The recipe identifies the date, encodes categorical calendar signals, removes zero-variance and linearly dependent columns, and normalizes numeric predictors.

```{r}
#| label: preprocessing
#| results: hide
bike_recipe <- recipe(rented_bike_count ~ ., data = training_data) |>
  update_role(date, new_role = "id") |>
  step_unknown(all_nominal_predictors()) |>
  step_novel(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_lincomb(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors())

linear_workflow <- workflow() |>
  add_recipe(bike_recipe) |>
  add_model(linear_reg() |> set_engine("lm"))

elastic_workflow <- workflow() |>
  add_recipe(bike_recipe) |>
  add_model(
    linear_reg(penalty = tune(), mixture = tune()) |>
      set_engine("glmnet")
  )

forest_workflow <- workflow() |>
  add_recipe(bike_recipe) |>
  add_model(
    rand_forest(mtry = tune(), min_n = tune(), trees = 300) |>
      set_engine(
        "ranger",
        importance = "permutation",
        num.threads = 1,
        seed = 2029
      ) |>
      set_mode("regression")
  )
```

Only the source code is shown here. The verbose internal printout of the recipe and workflow is intentionally suppressed because it adds no analytical value.

## Tuning and final fits

```{r}
#| label: tune-and-fit
#| results: hide
tuning_metrics <- metric_set(rmse, mae)

set.seed(2027)
elastic_grid <- grid_space_filling(
  penalty(range = c(-5, 0)),
  mixture(),
  size = 6
)

elastic_tuning <- tune_grid(
  elastic_workflow,
  resamples = temporal_resamples,
  grid = elastic_grid,
  metrics = tuning_metrics,
  control = control_grid(allow_par = FALSE)
)

best_elastic <- select_best(elastic_tuning, metric = "rmse")

set.seed(2028)
forest_grid <- grid_space_filling(
  mtry(range = c(4L, 18L)),
  min_n(range = c(5L, 40L)),
  size = 6
)

forest_tuning <- tune_grid(
  forest_workflow,
  resamples = temporal_resamples,
  grid = forest_grid,
  metrics = tuning_metrics,
  control = control_grid(allow_par = FALSE)
)

best_forest <- select_best(forest_tuning, metric = "rmse")

linear_fit <- fit(linear_workflow, training_data)
elastic_fit <- elastic_workflow |>
  finalize_workflow(best_elastic) |>
  fit(training_data)
set.seed(2029)
forest_fit <- forest_workflow |>
  finalize_workflow(best_forest) |>
  fit(training_data)
```

## Final holdout results

The benchmark averages historical demand by weekday and hour. If a combination is unavailable, it falls back to the hourly and then global training mean.

```{r}
#| label: predictions-and-metrics
calendar_average <- training_data |>
  summarise(calendar_average = mean(rented_bike_count), .by = c(hour, weekday))

hourly_average <- training_data |>
  summarise(hour_average = mean(rented_bike_count), .by = hour)

calendar_predictions <- testing_data |>
  left_join(calendar_average, by = c("hour", "weekday")) |>
  left_join(hourly_average, by = "hour") |>
  transmute(
    date,
    hour,
    rented_bike_count,
    .pred = coalesce(
      calendar_average,
      hour_average,
      mean(training_data$rented_bike_count)
    ),
    model = "Calendar baseline"
  )

linear_predictions <- testing_data |>
  select(date, hour, rented_bike_count) |>
  bind_cols(predict(linear_fit, new_data = testing_data)) |>
  mutate(.pred = pmax(0, .pred), model = "Linear regression")

elastic_predictions <- testing_data |>
  select(date, hour, rented_bike_count) |>
  bind_cols(predict(elastic_fit, new_data = testing_data)) |>
  mutate(.pred = pmax(0, .pred), model = "Elastic Net")

forest_predictions <- testing_data |>
  select(date, hour, rented_bike_count) |>
  bind_cols(predict(forest_fit, new_data = testing_data)) |>
  mutate(.pred = pmax(0, .pred), model = "Random forest")

predictions <- bind_rows(
  calendar_predictions,
  linear_predictions,
  elastic_predictions,
  forest_predictions
)

peak_threshold <- quantile(
  training_data$rented_bike_count,
  probs = 0.90,
  names = FALSE
)

model_metrics <- predictions |>
  summarise(
    rmse = sqrt(mean((rented_bike_count - .pred)^2)),
    mae = mean(abs(rented_bike_count - .pred)),
    rsq = rsq_trad_vec(rented_bike_count, .pred),
    wape_pct = 100 * sum(abs(rented_bike_count - .pred)) /
      sum(rented_bike_count),
    bias = mean(.pred - rented_bike_count),
    peak_mae = mean(
      abs(rented_bike_count[rented_bike_count >= peak_threshold] -
        .pred[rented_bike_count >= peak_threshold])
    ),
    .by = model
  ) |>
  arrange(rmse)

leader <- model_metrics |>
  slice_min(rmse, n = 1)

baseline_rmse <- model_metrics |>
  filter(model == "Calendar baseline") |>
  pull(rmse)

improvement <- 100 * (baseline_rmse - leader$rmse) / baseline_rmse

model_metrics |>
  mutate(
    rmse = round(rmse, 1),
    mae = round(mae, 1),
    rsq = round(rsq, 1),
    wape_pct = round(wape_pct, 1),
    bias = round(bias, 1),
    peak_mae = round(peak_mae, 1)
  ) |>
  kable(caption = "Operational metrics on unseen future dates")
```

::: {.metric-grid}
::: {.metric-card}
<strong>`r leader$model`</strong>
<span>Best final model</span>
:::
::: {.metric-card}
<strong>`r sprintf("%.1f", leader$rmse)` bikes</strong>
<span>Holdout RMSE</span>
:::
::: {.metric-card}
<strong>`r sprintf("%.1f%%", improvement)`</strong>
<span>RMSE gain vs calendar baseline</span>
:::
::: {.metric-card}
<strong>`r sprintf("%.0f", leader$peak_mae)` bikes</strong>
<span>Peak-demand MAE</span>
:::
:::

::: {.callout-important}
## Decision takeaway
Use the winning model as a planning signal, not an autonomous dispatch rule. It materially improves the calendar benchmark, but its negative bias and larger peak error call for an operational safety buffer.
:::

```{r}
#| label: model-comparison
#| fig-cap: "Lower RMSE is better; evaluation uses only the final future holdout."
model_metrics |>
  ggplot(aes(rmse, reorder(model, -rmse), fill = model)) +
  geom_col(show.legend = FALSE, width = 0.68) +
  geom_text(aes(label = round(rmse, 1)), hjust = -0.15, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.14))) +
  scale_fill_viridis_d(end = 0.85) +
  labs(title = "Final holdout error by model", x = "RMSE (bikes)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), panel.grid.major.y = element_blank())
```

## Operational diagnostics

```{r}
#| label: holdout-trace
#| fig-cap: "Observed demand and the winning model over the final seven days."
winning_model <- leader$model[[1]]

predictions |>
  filter(model == winning_model, date >= max(date) - 6) |>
  mutate(timestamp = as.POSIXct(date) + as.numeric(as.character(hour)) * 3600) |>
  select(timestamp, observed = rented_bike_count, predicted = .pred) |>
  pivot_longer(c(observed, predicted), names_to = "series", values_to = "bikes") |>
  ggplot(aes(timestamp, bikes, color = series)) +
  geom_line(linewidth = 0.75) +
  scale_color_manual(values = c(observed = "#264653", predicted = "#E76F51")) +
  labs(
    title = str_c("Observed and predicted demand — ", winning_model),
    x = NULL, y = "Bikes", color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```

```{r}
#| label: peak-errors
#| fig-cap: "Positive residuals indicate under-forecasting."
predictions |>
  filter(model == winning_model) |>
  mutate(
    hour = as.integer(as.character(hour)),
    residual = rented_bike_count - .pred,
    peak = rented_bike_count >= peak_threshold
  ) |>
  ggplot(aes(hour, residual, color = peak)) +
  geom_hline(yintercept = 0, color = "#6C757D") +
  geom_jitter(width = 0.18, alpha = 0.25) +
  geom_smooth(se = FALSE, color = "#E76F51") +
  scale_color_manual(values = c(`FALSE` = "#457B9D", `TRUE` = "#D62828")) +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(
    title = "Peak-hour errors remain the hardest problem",
    x = "Hour", y = "Observed minus predicted", color = "Peak demand"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```

```{r}
#| label: feature-importance
#| fig-cap: "Permutation importance describes predictive contribution, not causality."
variable_importance <- forest_fit |>
  extract_fit_engine() |>
  pluck("variable.importance") |>
  enframe(name = "variable", value = "importance") |>
  arrange(desc(importance)) |>
  slice_head(n = 12)

variable_importance |>
  ggplot(aes(importance, reorder(variable, importance))) +
  geom_col(fill = "#2A9D8F") +
  labs(title = "Random forest permutation importance", x = "Importance", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), panel.grid.major.y = element_blank())
```

Temperature, trend, moisture conditions, solar radiation, and commute-hour indicators carry much of the signal. This does not make them causal levers; it makes them useful forecasting inputs.

## From forecast to action

- Refresh the forecast before each operating shift with the latest weather outlook.
- Convert expected rentals into station-level targets only after adding station inventory and trip-flow data.
- Add a peak buffer when the cost of stockouts exceeds the cost of carrying idle bicycles.
- Monitor WAPE, bias, and peak MAE by season; retrain when performance drifts beyond agreed thresholds.
- Keep closures outside the demand model and handle them with an explicit service-status rule.

The dataset covers one year and has no station, capacity, pricing, event, or forecast-weather fields. The result demonstrates sound model evaluation, but it is not yet a production rebalancing system. See the [methodology](docs/methodology.md), [model card](docs/model-card.md), and [data dictionary](docs/data-dictionary.md) for the complete audit trail.
