---
title: "Predicting Seoul Bike Rental Demand"
subtitle: "Temporal holdout comparison with tidymodels"
author: "Miguelangel Palma Rojas"
date: last-modified
lang: en
---

```{r}
#| label: setup
#| include: false
source(here::here("R", "analysis.R"))
bikes <- read_bike_snapshot()
model_results <- fit_bike_models(bikes)
```

## Business objective

Hourly demand forecasts help operators position bicycles before peaks and reduce empty or saturated stations. This reconstruction uses the original, unnormalized rental count so model errors remain interpretable as bicycles.

## Data and leakage controls

The UCI snapshot contains `r nrow(bikes)` hourly records. Date-derived weekday and weekend features are created before modeling. Non-functioning periods are excluded because zero rentals during closure do not represent customer demand.

The earliest 80% of observations train and tune the models; the latest 20% form a final holdout. The date itself is excluded from the recipe, while valid calendar and weather information remains available.

```{r}
#| fig-alt: "Line chart of average bike rental demand by hour for each season."
plot_hourly_demand(bikes, "en")
```

## Models

- **Mean baseline** establishes the minimum benchmark.
- **Linear regression** estimates additive effects.
- **Elastic Net** tunes both regularization strength and the L1/L2 mixture.
- **Random forest** captures nonlinearities and interactions without manually creating polynomial terms.

Numeric predictors are normalized inside each workflow. Categorical variables are encoded through the recipe, preventing preprocessing from learning from the final test period.

## Final holdout performance

```{r}
model_results$metrics |>
  tidyr::pivot_wider(names_from = .metric, values_from = .estimate) |>
  arrange(rmse) |>
  knitr::kable(digits = 2)
```

RMSE and MAE are measured in rented bicycles. R² uses the traditional holdout definition relative to the test-set variation.

## What drives predictions?

```{r}
#| fig-alt: "Horizontal bars ranking the most important random forest predictors."
plot_variable_importance(model_results$importance, "en")
```

Permutation importance describes predictive contribution, not causal impact. Correlated weather and calendar variables can share importance.

## Operational interpretation

The model can support staffing and redistribution planning, but deployment would require recent observations, monitoring for seasonal drift, and station-level data. Citywide hourly totals cannot optimize individual station inventory by themselves.

## Conclusion

The rebuilt analysis removes the original train/test metric error, avoids redundant dummy variables, preserves real rental units, and compares transparent and nonlinear models on unseen future observations.

