---
title: "Predicting Seoul Bike Rental Demand"
subtitle: "Temporal holdout comparison with tidymodels"
author: "Miguelangel Palma Rojas"
date: last-modified
lang: en
---

## Business objective

Hourly forecasts help operators position bicycles before demand peaks. This notebook displays the full workflow: validation, feature typing, temporal splitting, preprocessing, tuning, holdout metrics, variable importance, and residual diagnostics.

## Load packages

```{r}
#| label: packages
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(here)
library(knitr)
library(readr)
library(tibble)
library(tidymodels)
library(tidyr)
```

## Read and validate the snapshot

```{r}
#| label: read-data
bikes <- read_csv(
  here("data", "seoul_bike_snapshot.csv"),
  show_col_types = FALSE,
  col_types = cols(date = col_date())
) |>
  mutate(
    weekday = factor(weekday),
    season = factor(season),
    holiday = factor(holiday),
    functioning_day = factor(functioning_day),
    is_weekend = factor(is_weekend)
  ) |>
  arrange(date, hour)

required_columns <- c(
  "date", "rented_bike_count", "hour", "temperature_c",
  "humidity_pct", "wind_speed_ms", "visibility_10m",
  "dew_point_c", "solar_radiation_mj_m2", "rainfall_mm",
  "snowfall_cm", "season", "holiday", "functioning_day",
  "weekday", "is_weekend"
)

missing_columns <- setdiff(required_columns, names(bikes))

if (length(missing_columns) > 0) {
  stop(
    "Bike data is missing columns: ",
    paste(missing_columns, collapse = ", ")
  )
}

stopifnot(
  !any(bikes$rented_bike_count < 0, na.rm = TRUE),
  anyDuplicated(bikes[c("date", "hour")]) == 0,
  all(bikes$hour %in% 0:23)
)

bikes |>
  summarise(
    observations = n(),
    first_date = min(date),
    last_date = max(date),
    average_rentals = mean(rented_bike_count),
    maximum_rentals = max(rented_bike_count)
  ) |>
  kable(digits = 1)
```

## Explore demand by hour and season

```{r}
#| label: hourly-demand-data
hourly_demand <- bikes |>
  filter(functioning_day == "Yes") |>
  group_by(hour, season) |>
  summarise(
    average_demand = mean(rented_bike_count),
    .groups = "drop"
  )
```

```{r}
#| label: hourly-demand-chart
#| fig-alt: "Line chart of average bike rental demand by hour for each season."
ggplot(
  hourly_demand,
  aes(x = hour, y = average_demand, color = season)
) +
  geom_line(linewidth = 1) +
  scale_color_viridis_d(end = 0.85) +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(
    title = "Demand changes by hour and season",
    x = "Hour",
    y = "Average rented bikes",
    color = "Season"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

## Define leakage-safe modeling data

Closure periods are excluded because their zero counts represent unavailable service, not customer demand. The raw date and redundant weekend flag are removed; weekday remains available.

```{r}
#| label: modeling-data
modeling_data <- bikes |>
  filter(functioning_day == "Yes") |>
  select(-functioning_day, -is_weekend)

set.seed(2026)

bike_split <- initial_time_split(modeling_data, prop = 0.8)
training_data <- training(bike_split)
testing_data <- testing(bike_split)
resamples <- vfold_cv(training_data, v = 3)

tibble(
  partition = c("Training and tuning", "Final test"),
  observations = c(nrow(training_data), nrow(testing_data)),
  first_date = c(min(training_data$date), min(testing_data$date)),
  last_date = c(max(training_data$date), max(testing_data$date))
) |>
  kable()
```

## Mean baseline

```{r}
#| label: baseline-model
baseline_predictions <- testing_data |>
  transmute(
    rented_bike_count,
    .pred = mean(training_data$rented_bike_count)
  )
```

## Preprocessing recipe

The date is excluded from the formula. Categorical levels are handled before dummy encoding, zero-variance columns are removed, and numeric normalization is estimated only from analysis folds.

```{r}
#| label: preprocessing-recipe
bike_recipe <- recipe(
  rented_bike_count ~
    hour +
    temperature_c +
    humidity_pct +
    wind_speed_ms +
    visibility_10m +
    dew_point_c +
    solar_radiation_mj_m2 +
    rainfall_mm +
    snowfall_cm +
    season +
    holiday +
    weekday,
  data = training_data
) |>
  step_unknown(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())
```

## Linear regression

```{r}
#| label: linear-model
linear_workflow <- workflow() |>
  add_recipe(bike_recipe) |>
  add_model(
    linear_reg() |>
      set_engine("lm")
  )

linear_fit <- fit(linear_workflow, data = training_data)

linear_predictions <- bind_cols(
  testing_data |> select(rented_bike_count),
  predict(linear_fit, new_data = testing_data)
)
```

## Elastic Net

```{r}
#| label: elastic-net-model
elastic_workflow <- workflow() |>
  add_recipe(bike_recipe) |>
  add_model(
    linear_reg(
      penalty = tune(),
      mixture = tune()
    ) |>
      set_engine("glmnet")
  )

elastic_grid <- grid_space_filling(
  penalty(range = c(-5, 1)),
  mixture(),
  size = 6
)

elastic_results <- tune_grid(
  elastic_workflow,
  resamples = resamples,
  grid = elastic_grid,
  metrics = metric_set(rmse, mae, rsq_trad)
)

best_elastic <- select_best(elastic_results, metric = "rmse")
best_elastic |> kable(digits = 5)

elastic_fit <- elastic_workflow |>
  finalize_workflow(best_elastic) |>
  fit(data = training_data)

elastic_predictions <- bind_cols(
  testing_data |> select(rented_bike_count),
  predict(elastic_fit, new_data = testing_data)
)
```

## Random forest

```{r}
#| label: random-forest-model
forest_workflow <- workflow() |>
  add_recipe(bike_recipe) |>
  add_model(
    rand_forest(
      mtry = tune(),
      min_n = tune(),
      trees = 300
    ) |>
      set_engine("ranger", importance = "permutation") |>
      set_mode("regression")
  )

forest_grid <- grid_space_filling(
  mtry(range = c(3L, 12L)),
  min_n(range = c(5L, 35L)),
  size = 6
)

forest_results <- tune_grid(
  forest_workflow,
  resamples = resamples,
  grid = forest_grid,
  metrics = metric_set(rmse, mae, rsq_trad)
)

best_forest <- select_best(forest_results, metric = "rmse")
best_forest |> kable()

forest_fit <- forest_workflow |>
  finalize_workflow(best_forest) |>
  fit(data = training_data)

forest_predictions <- bind_cols(
  testing_data |> select(rented_bike_count),
  predict(forest_fit, new_data = testing_data)
)
```

## Final holdout performance

```{r}
#| label: holdout-metrics
baseline_metrics <- bind_rows(
  rmse(baseline_predictions, rented_bike_count, .pred),
  mae(baseline_predictions, rented_bike_count, .pred),
  rsq_trad(baseline_predictions, rented_bike_count, .pred)
) |>
  mutate(model = "Mean baseline")

linear_metrics <- bind_rows(
  rmse(linear_predictions, rented_bike_count, .pred),
  mae(linear_predictions, rented_bike_count, .pred),
  rsq_trad(linear_predictions, rented_bike_count, .pred)
) |>
  mutate(model = "Linear regression")

elastic_metrics <- bind_rows(
  rmse(elastic_predictions, rented_bike_count, .pred),
  mae(elastic_predictions, rented_bike_count, .pred),
  rsq_trad(elastic_predictions, rented_bike_count, .pred)
) |>
  mutate(model = "Elastic Net")

forest_metrics <- bind_rows(
  rmse(forest_predictions, rented_bike_count, .pred),
  mae(forest_predictions, rented_bike_count, .pred),
  rsq_trad(forest_predictions, rented_bike_count, .pred)
) |>
  mutate(model = "Random forest")

holdout_metrics <- bind_rows(
  baseline_metrics,
  linear_metrics,
  elastic_metrics,
  forest_metrics
) |>
  select(model, .metric, .estimate)

holdout_metrics |>
  pivot_wider(
    names_from = .metric,
    values_from = .estimate
  ) |>
  arrange(rmse) |>
  kable(digits = 2)
```

RMSE and MAE remain interpretable as rented bicycles.

## Variable importance

```{r}
#| label: variable-importance-data
forest_engine <- extract_fit_engine(forest_fit)

variable_importance <- enframe(
  forest_engine$variable.importance,
  name = "variable",
  value = "importance"
) |>
  arrange(desc(importance)) |>
  slice_head(n = 12)
```

```{r}
#| label: variable-importance-chart
#| fig-alt: "Horizontal bars ranking the most important random forest predictors."
ggplot(
  variable_importance,
  aes(x = importance, y = reorder(variable, importance))
) +
  geom_col(fill = "#2A9D8F") +
  labs(
    title = "Random forest permutation importance",
    x = "Importance",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.major.y = element_blank()
  )
```

## Residual diagnosis

```{r}
#| label: residual-chart
#| fig-alt: "Residuals plotted against random forest predictions for the final test period."
forest_diagnostics <- forest_predictions |>
  mutate(residual = rented_bike_count - .pred)

ggplot(forest_diagnostics, aes(x = .pred, y = residual)) +
  geom_hline(yintercept = 0, color = "#555555") +
  geom_point(alpha = 0.35, color = "#2A9D8F") +
  geom_smooth(se = FALSE, color = "#E76F51") +
  labs(
    title = "Random forest residuals on unseen observations",
    x = "Predicted rented bikes",
    y = "Observed minus predicted"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Conclusion

The visible code demonstrates temporal evaluation, leakage controls, two tuning strategies, nonlinear modeling, interpretable metrics, importance, and residual analysis. See the [Spanish version](es/index.qmd).
